---
title: "Data Manipulation with `dplyr`"
output: github_document
---

```{r}
library(tidyverse)

options(tibble.print_min = 3) #only print first 3 lines of each tibble

#load litters data as tibbles and clean names
litters_data = read_csv("./data/data_import_examples/FAS_litters.csv")
litters_data = janitor::clean_names(litters_data)

#load pups data as tibbles and clean names
pups_data = read_csv("./data/data_import_examples/FAS_pups.csv")
pups_data = janitor::clean_names(pups_data)
```

### `select`
Extract a subset of cols(var) in a data table

```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive) #select(df, var1, var2, var3)
```
```{r}
select(litters_data, group:gd_of_birth) #select a range of cols
```
Specify cols to remove
```{r}
select(litters_data, -pups_survive)
```
Rename var
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
#rename+select, new_name = old_name
```

If only want to rename not select, can use `rename` instead of `select`.
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```
Use `?select_helpers` to see helper functions of  `select`, like `starts_with()`, `ends_with()`, `contains()`

```{r}
select(litters_data, starts_with("gd"))
```
`everything()` reorganize cols
```{r}
select(litters_data, litter_number, pups_survive, everything()) #pulls litter_number and pups_survive to front
```

`relocate` (not select helper) does similar thing
```{r}
relocate(litters_data, litter_number, pups_survive)
```

`select` export df not one column

### `filter`

select rows(observation), often filter using comparison operators(`>`, `>=`, `<`, `<=`, `==`, and `!=`).
`%in%` detects if values appear in a set.
`is.na()` finds missing values
`&` and, `|` or, `!` negate

Some ways you might filter the litters data are:
* `gd_of_birth == 20`
* `pups_survive != 4`
* `group %in% c("Con7", "Con8")` select rows where group(col name) is Con7 or Con8
* `group == "Con7" & gd_of_birth == 20`

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

To omit missing observation, use `drop_na` from `tidyr` package instead of using `filter`.
* `drop_na(litters_data)` remove any row with any missing value
* `drop_na(litters_data, wt_increase)` remove any row for which `wt_increase` is missing.

### `mutate`
Change or create new cols

```{r}
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       #create new column wt_gain
       group = str_to_lower(group)
       #change group variables to lower case
       )
```
* new var appear at the end of the dataset
* can overwrite old var
* can create new var and immediately refer to it

### `arrange`
arrange rows according to the values in one or more columns

```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
#sort litters_data on group first and pups_born_alive second, print first 10 lines
```
can also sort in descending order
```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)
#sort rows according to group descendingly first, pups_born_alive ascendingly second
```

### `%>%`

```{r}
litters_data =
  read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```
Ctrl + Shift + M shortcut for `%>%`

can use `.` as placeholder for the result of preceding call
```{r}
litters_data = 
  read_csv("./data/data_import_examples/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names(dat = .) %>%
  select(.data = ., -pups_survive) %>%
  mutate(.data = .,
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(data = ., wt_gain)
```


